VOI-aware MCTS

### 2012-02-26 paper/vago.ecai/paper.tex
short paper started

\documentclass{article}
\usepackage{enumerate}
\usepackage{algpseudocode}
\usepackage[ruled]{algorithm}
\usepackage{url}
\usepackage{framed}
\usepackage{amsfonts,amsmath,amsthm,amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{color}
\usepackage{geometry}

@@

\title{VOI-aware Monte-Carlo Sampling in Trees}
\author {David Tolpin, Solomon Eyal Shimony \\
Department of Computer Science, \\
Ben-Gurion University of the Negev, Beer Sheva, Israel \\
\{tolpin,shimony\}@cs.bgu.ac.il}

@@

\begin{abstract}
Upper bounds for the VOI are provided for pure exploration in the
Multi-armed Bandit Problem. Sampling policies based on the upper
bounds are suggested. Empirical evaluation of the policies and
comparison to the UCB1 and UCT policies is provided
on random problem instances as well as on the Go game.
\end{abstract}

@@

\section{Some Concentration Inequalities}

Let $X_1, \ldots, X_n$ be i.i.d. random variables with values from $[0,1]$,
$X=\frac 1 n \sum_{i=1}^n X_i$. Then 
\begin{description}
\item[Hoeffding inequality \rm{\cite{Hoeffding.ineq}}:] 
\begin{equation}
\Pr(X-\IE[X] \ge a) \le \exp ( -2a^2n)
\label{eqn:conc-hoeffding}
\end{equation}
\item[Empirical Bernstein inequality \rm{(derived in Appendix~\ref{app:deriv-conc-empbernstein})}:]
\begin{eqnarray}
\Pr(X-\IE[X] \ge a) &\le& 2\exp \left( - \frac {a^2n} {\frac {14} {3}
                          \frac {n} {n-1}a+2\overline\sigma_n^2}\right)\nonumber\\
                    &\le& 2\exp \left( - \frac {a^2n} {10a+2\overline\sigma_n^2}\right)
\label{eqn:conc-empbernstein}
\end{eqnarray}
where sample variance $\overline\sigma_n^2$ is
\begin{equation}
\overline\sigma_n^2=\frac 1 {n(n-1)} \sum_{1\le i < j\le n}(X_i-X_j)^2
\label{eqn:sample-variance}
\end{equation}
\end{description}
Bounds (\ref{eqn:conc-hoeffding}, \ref{eqn:conc-empbernstein}) are symmetrical
around the mean. Bound~(\ref{eqn:conc-empbernstein}) is tighter than
(\ref{eqn:conc-hoeffding}) for small $a$ and $\overline\sigma_n^2$.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.8]{flat.pdf}
\caption{Random instances: regret vs. number of samples}
\label{fig:random-instances}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.8]{vct-wins.pdf}
\caption{Go: winning rate --- UCT against VCT}
\label{fig:uct-against-vct}
\end{figure}
\begin{figure}[t]
\centering
\includegraphics[scale=0.8]{ect-wins.pdf}
\caption{Go: winning rate --- UCT against ECT}
\label{fig:uct-against-ect}
\end{figure}
\begin{figure}[t]
\centering
\includegraphics[scale=0.8]{bct-wins.pdf}
\caption{Go: winning rate --- UCT against BCT}
\label{fig:uct-against-bct}
\end{figure}
The engines were compared on the 9x9 board, for 5000, 7000, 1000, and
15000 samples (game simulations) per ply, each experiment was repeated
1000 times. Figures~\ref{fig:uct-against-vct}--\ref{fig:uct-against-bct}
show the winning rate of UCT against VCT, ECT, BCT, correspondingly,
vs. the stopping threshold (if the maximum VOI of a sample is below
the threshold, the simulation is stopped, and a move is chosen). Each
curve in the figures corresponds to a certain number of samples per
ply. The lower the curve, the better is the corresponding policy
compared to UCT.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.8]{bests-bw.pdf}
\caption{Go: best winning rate comparison}
\label{fig:best-winning-rate}
\end{figure}
Figure~\ref{fig:best-winning-rate} presents relative performance of
different VOI-aware policies depending on the number of samples per
ply by comparing the best achieved winning rates against UCT (again,
the lower the bar, the better is the policy). Both ECT (based on the
Hoeffding inequality with an intermediate point) and BCT (based on the empirical
Bernstein inequality) are better than VCT. The difference between ECT
and VCT is most prominent for smaller number of samples, and the
winning rate of ECT increases relatively slowly with the number of
samples. In contrast, BCT is worse than ECT and only slightly better
than VCT for smaller number of samples, but the winning rate rapidly
increases with the number of samples, and for 10000 and 15000 samples
per ply BCT is the best of the policies with the winning rate reaching
67\%.

@@

\section{Empirical Bernstein Inequality}
\label{app:deriv-conc-empbernstein}

Theorem~4 in~\cite{MaurerPontil.benrstein} states that
\begin{equation}
\Pr\left(\IE[X]-\overline X_n
    \ge \sqrt { \frac {2\overline\sigma_n^2 \ln 2/\delta} n } + \frac {7 \ln 2/\delta} {3(n-1)}\right)\le \delta
\end{equation}
Therefore
\begin{equation}
\Pr\left(\IE[X]-\overline X_n \ge
              \sqrt { \left(\frac {7 \ln 2/\delta} {3(n-1)}\right)^2+\frac {2\overline\sigma_n^2 \ln 2/\delta} n }
                      + \frac {7 \ln 2/\delta} {3(n-1)}\right)
   \le \delta.
\end{equation}
$a=\sqrt { \left(\frac {7 \ln 2/\delta} {3(n-1)}\right)^2+\frac {2\overline\sigma_n^2 \ln 2/\delta} n } + \frac {7 \ln 2/\delta} {3(n-1)}$
 is a root of square equation $a^2-a\frac {14 \ln 2/\delta} {3(n-1)} -\frac {2\overline\sigma_n^2 \ln 2/\delta} n=0$
which, solved for $\delta\triangleq\Pr(\IE[X]-\overline X_n\ge a)$,
gives
\begin{equation*}
\Pr(\IE[X]-\overline X_n\ge a)\le 2\exp \left( - \frac {na^2} {\frac {14} {3} \frac {n} {n-1}a+2\overline\sigma_n^2}\right)
\end{equation*}
Other derivations, giving slightly different results, are possible.

\section{Better Hoeffding-Based Bound on Value of Perfect Information}
\label{app:better-hoeffding-bound}

Bounds (\ref{eqn:bound-perf-hoeffding}, \ref{eqn:bound-blnk-hoeffding}) have the form
\begin{equation}
  \Lambda_i \le \hat\Lambda_i=\left\{
  \begin{array}{l l}
    (\overline X_\beta-B)\exp\left(- \gamma(\overline X_i - \overline X_\beta)^2 n_i \right) & \mbox{{\rm if} $i=\alpha$} \\
    (A-\overline  X_\alpha)\exp\left(- \gamma(\overline X_\alpha - \overline X_i)^2 n_i \right)  &  \mbox{\rm otherwise}
  \end{array} \right.
\label{eqn:bound-generic-hoeffding}
\end{equation}
where $A$ is the upper bound on the posterior sample mean $\overline
X_i'$ if $i\ne \alpha$, $B$ is the lower bound on $\overline X_i'$ if
$i=\alpha$. For $i=\alpha$, the bound can be improved by selecting an
intermediate point $y$, $B < y < \overline X_\beta$, and computing the bound
as the sum of two parts: 
\begin{itemize}
\item $\overline X_\beta-y$ multiplied by the probability that
  $\overline X_i' \le \overline X_\beta$;
\item $\overline X_\beta-B$ multiplied by the probability that $\overline X_i'\le
  y$.
\end{itemize}.
\begin{equation}
\Lambda_{i|i=\alpha} \le \hat \Lambda_{i|i=\alpha} =
 (\overline X_\beta-y)\exp\left(-\gamma (\overline X_i-\overline X_\beta)^2n_i\right)
 +(\overline X_\beta-B)\exp\left(-\gamma (\overline X_i-y)^2n_i\right)
\label{eqn:bound-twopart-hoeffding-alpha}
\end{equation}
If $y$ that minimizes $\hat\Lambda_{i|i=\alpha}$ exists, the minimum is achieved when
 $\frac {d\hat\Lambda_{i|i=\alpha}} {dy}=0$, that
is, when $y$ is the root of the following equation:
\begin{equation}
  2\gamma (\overline X_\beta-B)(\overline X_i-y)n_i = \exp\left(-\gamma \left(\frac {\overline X_i-\overline X_\beta}
    {\overline X_i-y}\right)^2n_i\right)
\end{equation}
The derivation for the case $i\ne \alpha$ is obtained by substitution $1-A, 1-\overline X_i, 1-\overline X_\alpha, 1-y$ instead of
$B, \overline X_i, \overline X_\beta, y$ into (\ref{eqn:bound-twopart-hoeffding-alpha}):
\begin{equation}
\Lambda_{i|i\ne\alpha} \le \hat\Lambda_{i|i\ne\alpha} = (y-\overline X_\alpha)\exp\left(-\gamma(\overline X_\alpha-\overline
  X_i)^2 n_i\right)+(A-\overline X_\alpha)\exp\left(-\gamma (y-\overline X_i)^2n_i\right)
\label{eqn:bound-twopart-hoeffding-rest}
\end{equation}
A closed-form solution for $y$ cannot be obtained, but given
$A, B, \overline X_\alpha, \overline X_\beta, n$, the value of $y$ can be efficiently
computed.

\bibliographystyle{plain}
\bibliography{refs}

>>>>

\documentclass{ecai2012}
\usepackage{enumerate}
\usepackage{algpseudocode}
\usepackage[ruled]{algorithm}
\usepackage{url}
\usepackage{framed}
\usepackage{amsfonts,amsmath,amsthm,amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{color}
\usepackage{geometry}

@@

\title{VOI-aware MCTS}
\author {David Tolpin, Solomon Eyal Shimony \\
Department of Computer Science, \\
Ben-Gurion University of the Negev, Beer Sheva, Israel \\
\{tolpin,shimony\}@cs.bgu.ac.il}

@@

\begin{abstract}
Upper bounds for the VOI are provided for pure exploration in the
Multi-armed Bandit Problem. A sampling policy based on the upper
bounds is suggested. Empirical evaluation of the policy and
comparison to the UCB1 and UCT policies is provided
on random problem instances as well as on the Go game.
\end{abstract}

@@

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{flat.pdf}
\caption{Random instances: regret vs. number of samples}
\label{fig:random-instances}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{vct-wins.pdf}
\caption{Go: winning rate --- UCT against VCT}
\label{fig:uct-against-vct}
\end{figure}
\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{ect-wins.pdf}
\caption{Go: winning rate --- UCT against ECT}
\label{fig:uct-against-ect}
\end{figure}
\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{bct-wins.pdf}
\caption{Go: winning rate --- UCT against BCT}
\label{fig:uct-against-bct}
\end{figure}
The engines were compared on the 9x9 board, for 5000, 7000, 1000, and
15000 samples (game simulations) per ply, each experiment was repeated
1000 times. Figures~\ref{fig:uct-against-vct}--\ref{fig:uct-against-bct}
show the winning rate of UCT against VCT, ECT, BCT, correspondingly,
vs. the stopping threshold (if the maximum VOI of a sample is below
the threshold, the simulation is stopped, and a move is chosen). Each
curve in the figures corresponds to a certain number of samples per
ply. The lower the curve, the better is the corresponding policy
compared to UCT.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.5]{bests-bw.pdf}
\caption{Go: best winning rate comparison}
\label{fig:best-winning-rate}
\end{figure}
Figure~\ref{fig:best-winning-rate} presents relative performance of
different VOI-aware policies depending on the number of samples per
ply by comparing the best achieved winning rates against UCT (again,
the lower the bar, the better is the policy). Both ECT (based on the
Hoeffding inequality with an intermediate point) and BCT (based on the empirical
Bernstein inequality) are better than VCT. The difference between ECT
and VCT is most prominent for smaller number of samples, and the
winning rate of ECT increases relatively slowly with the number of
samples. In contrast, BCT is worse than ECT and only slightly better
than VCT for smaller number of samples, but the winning rate rapidly
increases with the number of samples, and for 10000 and 15000 samples
per ply BCT is the best of the policies with the winning rate reaching
67\%.

@@

\bibliographystyle{ecai2012}
\bibliography{refs}
